---
title: "R Exercises for the OODA Short Course"
subtitle: "IBC, Seoul, Korea"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# R Exercise 1: Spanish Male Mortality data

Load the data for this exercise. Check that any external files are in your working directory before evaluating the chunk.

*Slide context:* R Exercise 1: Spanish Male Mortality data

```{r chunk 1}
x <- read.csv("Spain-Mx_1x1.csv")
x$Age[x$Age=="110+"]<-"110"
Age <- as.double(x$Age) ; Year <- as.double(x$Year)
Male <- as.double(x$Male) ; Female <- as.double(x$Female)
n<-114 ; d<-99
X <- matrix(0,n,d)
for (i in 1908:2021){
  for (j in 0:98){
 X[i-1907,j+1] <- Male[ (Age==j)&(Year==i)]   
  }
}
Y <- log10(X)
palette(rainbow(n+10))
plot( Y[1,], type="n",xlab="Age",ylab="Mortality rate",ylim=c(-4.5,0))
for ( i in 1:n){
  lines(x$Age[1:d],Y[i,],col=n+1-i)
}
legend("bottomright",c("1908","1918","1936","1960","1980","2000","2021"),
               col=c(n,n-10,n-28,n-52,n-72,n-92,n-113),pch=16,cex=0.5)
```

# R Exercise 1: Residuals

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 1: Residuals

```{r chunk 2}
Ybar <- apply( Y , 2 , mean)
plot( Ybar, type="l",xlab="Age",ylab="Mortality rate",ylim=c(-3.5,0))
palette(rainbow(n+10))
plot( Y[1,], type="n",xlab="Age",ylab="Mortality rate",ylim=c(-2,2))
for ( i in 1:n){
  lines(x$Age[1:d],Y[i,]-Ybar,col=n+1-i)
}
legend("bottomright",c("1908","1918","1936","1960","1980","2000","2021"),
          col=c(n,n-10,n-28,n-52,n-72,n-92,n-113),pch=16,cex=0.5)
```

# R Exercise 1: Mortality PCA

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 1: Mortality PCA

```{r chunk 3}
pca <- prcomp( Y )
plot(pca$center)
print(pca$sdev[1:2]**2/sum(pca$sdev**2)*100)
plot(pca$x[,1],pca$x[,2],xlab="PC1",ylab="PC2",col=n:1)
lines(pca$x[,1],pca$x[,2])
legend("bottomright",c("1908","1918","1936","1960","1980","2000","2021"),
                col=c(n,n-10,n-28,n-52,n-72,n-92,n-113),pch=16,cex=0.5)
plot( Y[1,], type="n",xlab="Age",ylab="Mortality",ylim=c(-2,2))
for ( i in 1:n){
  lines(x$Age[1:d],pca$x[i,1]*pca$rotation[,1],col=n+1-i)
}
legend("bottomright",c("1908","1918","1936","1960","1980","2000","2021"),
                 col=c(n,n-10,n-28,n-52,n-72,n-92,n-113),pch=16,cex=0.5)
plot( Y[1,], type="n",xlab="Age",ylab="Mortality",ylim=c(-0.5,0.5))
for ( i in 1:n){
  lines(x$Age[1:d],pca$x[i,2]*pca$rotation[,2],col=n+1-i)
}
legend("bottomright",c("1908","1918","1936","1960","1980","2000","2021"),
                   col=c(n,n-10,n-28,n-52,n-72,n-92,n-113),pch=16,cex=0.5)
```

# R Exercise 2: Pan Cancer data in R

Load the data for this exercise. Check that any external files are in your working directory before evaluating the chunk.

*Slide context:* The cancer groups are Bladder (BLCA), Kidney Renal (KIRC), Ovarian (OV), Head and Neck Squamous Cell (HNSC), Colon Adenocarcinoma (COAD), and Breast (BRCA). Pan Cancer data

```{r chunk 4}
palette("default")
x<-read.csv("PanCancer-ReducedGeneExpressionData.csv",header=TRUE)
dim(x)
w<-x[2:12479,2:301] # remove first row and column
w<-sapply(w,as.numeric)
group<-unname(x[1,2:301])
group<-c(as.matrix(group))
table(group)
gcol<-rep(1:6,each=50) # give each group a different color
X<-t(w)
dim(X)
pca <- prcomp(X)
plot(pca$sdev**2/sum(pca$sdev**2))
```

# R Exercise 2: Pan Cancer (continued)

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 2: Pan Cancer (continued) What do PC1 and PC2 represent?

```{r chunk 5}
plot(pca$x[,1],pca$x[,2],col=gcol,xlab="PC1",ylab="PC2",type="n")
text(pca$x[,1],pca$x[,2],group,col=gcol,cex=0.5)
plot(pca$rotation[,1])
plot(pca$rotation[,2])
plot(apply(X,2,sd),ylab="sd")
```

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* Carry out PCA with correlation matrix R

```{r chunk 6}
pca2 <- prcomp(X,scale=TRUE)
plot(pca2$sdev**2/sum(pca$sdev**2),ylab="percent")
plot(pca2$x[,1],pca2$x[,2],col=gcol,xlab="PC1",ylab="PC2",type="n")
text(pca2$x[,1],pca2$x[,2],group,col=gcol,cex=0.5)
```

# R Exercise 3: Fisher-Rao growth velocity data

Load the data for this exercise. Check that any external files are in your working directory before evaluating the chunk.

*Slide context:* R Exercise 3: Fisher-Rao growth velocity data Now consider FDA with Fisher-Rao registration. Consider the functional PCA with phase-warping of the growth velocity curve data in the fdasrvf library.

```{r chunk 7}
library("fdasrvf")
data(growth_vel)
n<-93
palette(rainbow(n+10))
plot(growth_vel$time,growth_vel$f[,1],ylim=c(-2,30),type="n",ylab="f")
for (i in 1:n){
lines(growth_vel$time,growth_vel$f[,i],col=n+10-i)
}
out <- time_warping( growth_vel$f, growth_vel$time)
plot(growth_vel$time,growth_vel$f[,1],ylim=c(-2,30),type="n",ylab="f")
for (i in 1:n){
lines(growth_vel$time,out$fn[,i],col=n+10-i)
}
```

# R Exercise 3: Fisher-Rao growth velocity (continued)

Carry out a principal component analysis and inspect the dominant modes of variation.

*Slide context:* R Exercise 3: Fisher-Rao growth velocity (continued) Three types of PCA can be explored:\\ horizontal: phase/warp PCA\\ vertical: amplitude PCA\\ joint: both phase and amplitude PCA\\ Fisher-Rao FPCA

```{r chunk 8}
hfpca <- horizFPCA(out, no = 3)
vfpca <- vertFPCA(out, no = 3)
jfpca <- jointFPCA(out, no = 3)
```

# R Exercise 4: novels as networks, top 100 words

Load the data for this exercise. Check that any external files are in your working directory before evaluating the chunk.

*Slide context:* R Exercise 4: novels as networks, top 100 words 16 Charles Dickens novels (1:16) and 7 Jane Austen novels (17:23). Select the top 100 words. Graph Laplacian

```{r chunk 9}
load("Austen-Dickens.RData")
library(GGally) ; library(network) ; library(sna) ; library(r.jive) ; 
library(shapes) ; library(intergraph) ; library(igraph)
range<-1:23
nr<-max(range)
novels <- 1:nr
ng1<-1
ng2<-100
ng<-ng2-ng1+1
L <- array(0, c(ng,ng,nr) )
words<-lookup[ng1:ng2]
words
```

# R Exercise 4: obtain graph Laplacians matrix and plot networks

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 4: obtain graph Laplacians matrix and plot networks

```{r chunk 10}
for (ii in 1:nr){
  print(ii)
  tem<- as.matrix( gl_wght_list[[ii]][ng1:ng2,ng1:ng2])
  A<-  diag(diag(tem)) -tem  #adjacency
  ones<-rep(1,times=ng)
  nodes<- c(A%*%ones)
  tot<-sum(nodes)
  GL<- (diag(nodes)-A)/tot  #scaled Graph Laplacian
   q90<-quantile(c(A),0.95)
  A[A<q90]<-0
  L[,,ii]<-GL
  ggn<-network(diag(A)-A)
}
  In<-asIgraph( ggn ) #plot the last network
  l2<-layout_in_circle(In)
  l <- layout_with_fr(In)
  plot(In,vertex.label=words,layout=l2,edge.arrow.mode=0,vertex.size=2,edge.curved=0.2)
  plot(ggnet2(ggn,edge.alpha=0.5,label=words))
  show.image(L[,,ii])
```

# R Exercise 4: novels as networks, top 100 words

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 4: novels as networks, top 100 words Compute distances, and MDS

```{r chunk 11}
dmat <- matrix(0,nr,nr)
for (i in (1:(nr-1))){
  for (j in (2:nr)){
    dmat[i,j]<- distcov( L[,,i], L[,,j] , method="Power", alpha=1/2)
    dmat[j,i]<-dmat[i,j]
  }
}
palette("default")
gpc<-c(rep(2,times=16),rep(4,times=7))
mds <- cmdscale(as.dist(dmat))
plot( mds , xlab="MDS1",ylab="MDS2", type="n")
text( mds, labels=novels , col=gpc)
legend("bottomleft",legend=c("Dickens","Austen"),pch=c(1,1),col=c(2,4))
```

# R Exercise 5: Pairwise Procrustes Fit

Compute distances or an ordination, then use the output to represent relationships between observations.

*Slide context:* R Exercise 5: Pairwise Procrustes Fit

```{r chunk 15}
centre <- function(X) scale(X, scale = FALSE)
csize  <- function(X) sqrt(sum(centre(X)^2))
preshape <- function(X) centre(X) / csize(X)
X <- dna.dat[,,1]
Y <- dna.dat[,,30]
X0 <- preshape(X)
Y0 <- preshape(Y)
sv <- svd(t(X0) %*% Y0)
R  <- sv$v %*% t(sv$u)
R
Y_aligned <- Y0 %*% R
dP <- sqrt(sum((X0 - Y_aligned)^2))
dP
print( sqrt( 2*(1-cos( riemdist( X,Y) ) )))   #check
print( riemdist( X,Y) )                           # similar
```

# R Exercise 5: Pairwise Procrustes Fit (continued)

Run this chunk as the next step in the exercise. The code is kept close to the original course material, with explanatory text added around it.

*Slide context:* R Exercise 5: Pairwise Procrustes Fit (continued) Carry out ordinary Procrustes registration on X,Y, with and without scaling.

```{r chunk 16}
X <- dna.dat[,,1]
Y <- dna.dat[,,30]
outline=c(1:22)
shapes3d( abind( X, Y) , joinline=outline )

#match Y to X, without scaling
opa1 <- procOPA( X, Y, scale=FALSE)
opa1
shapes3d( abind( opa1$Ahat , opa1$Bhat) , joinline=outline )

#match Y to X, with scaling
opa2 <- procOPA( X, Y, scale=TRUE) 
opa2
shapes3d( abind( opa2$Ahat , opa2$Bhat) , joinline=outline )
```

# R Exercise 6: GPA with the shapes package

Load the data for this exercise. Check that any external files are in your working directory before evaluating the chunk.

*Slide context:* R Exercise 6: GPA with the shapes package T2 vertebrae outlines

```{r chunk 17}
data(mice)
outline=c(1:60,1)
plotshapes(mice$outlines,joinline=outline)
# mice$outlines is k x m x n landmark array
gpa <- procGPA(mice$outlines)
aligned <- gpa$rotated.  # aligned configurations
mean_shape <- gpa$mshape   # mean shape

plotshapes(gpa$rotated,joinline=outline)
lines( mean_shape[c(1:60,1),], col=2)
```

# R Exercise 7: Two Scale Curves Data

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 7: Two Scale Curves Data Read in the data in TwoScaleCurvesData.xlsx and plot the curves. There are n=200 piecewise linear curves in d=100 dimensions. What structure do you see in the data objects?

```{r chunk 18}
library("readxl")
x <- read_xlsx("TwoScaleCurvesData.xlsx",col_names=FALSE)
y<-as.matrix(t(x))
n<-dim(y)[1]
y <- y[order(y[,1]),]
palette(rainbow(n+10))
plot( y[1,], type="n",ylim=c(min(y),max(y)),ylab="y")
for ( i in 1:n){
  lines(1:100,y[i,],col=n+1-i)
}
```

# R Exercise 7: PCA: Two Scale Curves Data.

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 7: PCA: Two Scale Curves Data. Now carry out PCA of the data. In particular obtain a scree plot and plot the first two PC scores. Do you see any patterns of interest?

```{r chunk 19}
pca <- prcomp(y)
percent <- 100*pca$sdev**2/sum(pca$sdev**2)
plot(percent)
lines(percent)
#PC scores
plot( pca$x[,1], pca$x[,2], col=1:n)
```

# R Exercise 7: Modes of variation: Two Scale Curves Data.

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 7: Modes of variation: Two Scale Curves Data. Investigate the first three modes of variation, and describe what you see.

```{r chunk 20}
plot(pca$rotation[,1]*pca$x[i,1],type="n",ylab="PC1 loadings x score",ylim=c(-60,60))
for (i in 1:n){
modeofvar1 <- pca$rotation[,1]*pca$x[i,1] 
  lines(1:100,modeofvar1,col=n+1-i)
}
plot(pca$rotation[,1]*pca$x[i,1],type="n",ylab="PC2 loadings x score",ylim=c(-60,60))
for (i in 1:n){
modeofvar2 <- pca$rotation[,2]*pca$x[i,2] 
  lines(1:100,modeofvar2,col=n+1-i)
}
plot(pca$rotation[,1]*pca$x[i,1],type="n",ylab="PC3 loadings x score",ylim=c(-60,60))
for (i in 1:n){
modeofvar3 <- pca$rotation[,3]*pca$x[i,3] 
  lines(1:100,modeofvar3,col=n+1-i)
}
```

# R Exercise 8: Drug Discovery Data

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 8: Drug Discovery Data Read in the Drug Discovery Data. There are two groups given in the first column of the spreadsheet. Carry out PCA.

```{r chunk 21}
palette("default")
x <- read_xlsx("DrugDiscoveryData.xlsx")
x <- as.matrix(x)
group <- as.double(x[,1])+1
xn <- as.matrix(x[,c(-1,-2,-3)])
n<-dim(xn)[1]
pca<-prcomp(xn)
percent <- 100*pca$sdev**2/sum(pca$sdev**2)
plot(percent)
lines(percent)
plot( pca$x[,1], pca$x[,2], col=group)
pairs( pca$x[,1:4] ,col=group)
```

# R Exercise 8: Standardizing: Drug Discovery Data

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 8: Standardizing: Drug Discovery Data Do you think the data should be standardized here? We now try standardizing, but we need to remove variables with no variability.

```{r chunk 22}
sdev <- apply( xn , 2, sd)
xn2 <- xn[,sdev!=0]
sd2 <- sdev[sdev!=0]
center <- apply( xn2, 2, mean)
dim(xn2)
ones<- rep(1,times=n)
xn3 <- (xn2 - ones%*%t(center)) %*%diag(1/sd2)
pca2<-prcomp(xn3,scale=TRUE)
percent <- 100*pca2$sdev**2/sum(pca2$sdev**2)
plot(percent)
lines(percent)
plot( pca2$x[,1], pca2$x[,2], col=group)
pairs( pca2$x[,1:4],col=group)
```

# R Exercise 8: Distance Weighted Discrimination (DWD). Drug Discovery Data.

Create the first visual checks of the data or fitted objects. Use these plots to understand the structure before moving on to summaries.

*Slide context:* R Exercise 8: Distance Weighted Discrimination (DWD). Drug Discovery Data. We now carry out DWD classification.

```{r chunk 23}
library(kerndwd) ; library(shapes)
Drug <- list()
#standardized X
Drug$X <- xn3
Drug$y <- as.factor( group*2-1 )
lambda = 10^(seq(-3, 3, length.out=10)) 
kern = vanilladot() 
m1 = kerndwd( Drug$X , Drug$y, kern, qval=1, lambda=lambda, eps=1e-5, maxit=1e5) 
plot(m1)
predict(m1, kern, Drug$X, tail(Drug$X))
aa<-matrix(m1$alpha[2:(n+1),1],1,n)
print(lambda[1])
DWDvec <- c(aa%*%Drug$X)
DWDvec <- DWDvec/Enorm(DWDvec)
pickbig <- order( -abs(DWDvec) )[1:25]
plot(sort(DWDvec))
properties <- colnames( xn2 )
properties[pickbig]
```

# R Exercise 8: DiProPerm Test: Drug Discovery Data.

Use resampling or permutation ideas to assess the stability or significance of the observed result.

*Slide context:* R Exercise 8: DiProPerm Test: Drug Discovery Data. Is there a significant difference between the two groups or not? Carry out a DiProPerm test.

```{r chunk 24}
B <- 200  ;  tvalues <- rep(0,times= B )
for (i in 1:B){ cat(c(i," ")) 
if (i==1){   Drug$X <- xn3 }
if (i > 1){ Drug$X <- xn3[sample(1:n),] }
Drug$y <- as.factor( group*2-1 )
lambda = 10^(seq(-3, 3, length.out=10)) 
kern = vanilladot() 
m1 = kerndwd( Drug$X , Drug$y, kern, qval=1, lambda=lambda, eps=1e-5, maxit=1e5) 
aa<-matrix(m1$alpha[2:(n+1),1],1,n)
DWDvec <- c(aa%*%Drug$X)
DWDvec <- DWDvec/Enorm(DWDvec)
DWDscore <- Drug$X%*%DWDvec
out <- t.test( DWDscore ~ group )
tvalues[i]<-abs ( out$statistic )   }
obs <- tvalues[1]  #observed t-statistic
hist(tvalues,nclass=30)   #permutation distribution, with observed value marked in red
lines( c(obs,obs) , c(0,20), col=2)   
print( (sum( tvalues > obs)+1)/(B) )  #pvalue
```

# R Example 9: Overlapping Classes Data

Carry out a principal component analysis and inspect the dominant modes of variation.

*Slide context:* R Example 9: Overlapping Classes Data Carry out PCA of the OverlappingClassesData.xlsx . PCA

```{r chunk 25}
x <- read_xlsx("OverlappingClassesData.xlsx",col_names=FALSE)
y<-as.matrix(t(x))
group<-  c(rep(1,times=50),rep(2,times=50))
n<-100
xn4 <- y 
pca<-prcomp(xn4)
plot(pca$x,col=group)
```
